---
title: "Regression and report generation"
author: "Niels Richard Hansen"
date: "February 4, 2016"
output:
  html_document:
    css: custom.css
    fig_height: 5
    fig_width: 5
    theme: flatly
---

# A practical exercise on using knitr

Solve this exercise using Markdown and knitr. Use 
[the code](https://nielsrhansen.github.io/courses/regression2016/Exercises/SAT.Rmd)
that generated this document as a template. 

From RStudio just click on 'Knit HTML' to get the result. Alternatively, you can 
run the following code. 


```{r knit, eval=FALSE}
library(knitr)
library(markdown)
knit("SAT.Rmd")   ### Creates the Markdown file
markdownToHTML("SAT.md", "SAT.html") ### Creates the html file
browseURL(paste('file://', file.path(getwd(), 'SAT.html'), sep='')) 
```

You can also read more about [using R and Markdown with RStudio](http://rmarkdown.rstudio.com).

The data set to consider in this exercise consists of mean verbal SAT scores for each of the 50 US states 
together with the percentage of students taking the test. 

To help you get started we read in the data and construct a plot. 

```{r readingData}
sat <- read.table(
  "https://nielsrhansen.github.io/courses/regression2016/Data/SAT.txt", 
  header = TRUE)
```


```{r scatterPlot, fig.width=6, fig.height=6, warning=FALSE}
library(tidyverse)
library(conflicted)
theme_set(theme_bw()) ## My choice of theme over the default. 
qplot(percent, verbalScore, data = sat)
```

## Question 1
*Fit a linear regression model with an intercept of the state SAT score as a function of percentage 
of students. Add the fitted line to the plot.*

Let us take a short look at the data:
```{r}
sat %>% head()
sat %>% summary()
(p_sat <- ggplot(sat, mapping = aes(x = percent, y = verbalScore)) + geom_point())
```
We make the basic linear model
```{r}
lm_basic <- lm(verbalScore ~ percent, data = sat)
```

We can also look at the result of the model
```{r}
summary(lm_basic)
```
We can plot the model as well, either directly from this model that we have just built, or via `ggplot`
```{r}
p_sat + geom_smooth(method = "lm")
```



## Question 2

Fit a piecewise affine function with the "bend" at 50% of the 
students taking the test. How can you formally test the linearity assumption in Question 1 
within this larger model?

*Hint: What you want to achieve in this question is to fit a linear model 
$$\mathbf{Y} = \mathbf{X} \beta + \varepsilon$$
where $\mathbf{X}$ is a matrix with 3 columns representing an intercept and 
two piecewise linear functions. You can, for instance, generate a basis using an
`ifelse` construction. Try different bases. Is there anyone particularly suitable for 
testing the linearity assumption?*

We may mutate the data frame to create an effect which activates post `percent>=50`. We could also do without this step and simply include this test in the formula directly.

We assume the linear model
$$
Y_i = X_i^T\beta + \varepsilon_i 
$$
with the model matrix $X$ parametrised such that it has four columns: One for the (baseline, `percent<50`) intercept, one for percent, one for the indicator of `percent>=50` and one for the `percent>=50` modulated slope. This could be an interaction model with two group levels, as to which half of `percent` (written as $p$ for short in the equation below, along with the indicator $ph$ (percent half)) the response value is for:
\begin{equation}
E(Y_i|X_{i,p},ph_{q(i)}) = \beta_0 + \beta_{p}X_{i,p}\\
+ (\beta_{ph_{\geq 50}} + \beta_{p, ph_{\geq 50}}X_{i,p})X_{i,ph_{\geq 50}},
\end{equation}
with $q(i)$ signifying allegiance to `percent<50` or `percent>=50` depending on whether $X_{i,p}$ is less than, or greater than or equal to $50.$

We may start by showcasing the fitting of such a linear model, via construction of the model matrix.
```{r}
X_names <- c("(Intercept)", "percent", "above50", "percent:above50")
X <- cbind(rep(1L,nrow(sat)), as.integer(sat$percent), sat$percent >= 50, as.integer(sat$percent * (sat$percent >= 50)))
colnames(X) <- X_names
X[sample((1:nrow(sat)), 6), ]
```

```{r}
lm_above50_matrix_lmfit <- lm.fit(x = X, y = sat$verbalScore)
lm_above50_matrix_lm <- lm(verbalScore ~ X + 0, data = sat)
lm_above50_matrix_lmfit$coefficients
```

```{r}
summary(lm_above50_matrix_lm)
```
We may also implement the traditional `lm`-call
```{r}
lm_above50 <- lm(verbalScore ~ percent * I(percent>=50), data = sat)
summary(lm_above50)
```
Using the estimates we may draw the resulting model:
```{r}
lm_above50_coef <- lm_above50_matrix_lmfit$coefficients
lm_above50_func <- function(percent) {
  return(lm_above50_coef[["(Intercept)"]] + lm_above50_coef[["percent"]]*percent + (lm_above50_coef[["above50"]]+lm_above50_coef[["percent:above50"]]*percent)*(percent>=50))
}

p_sat + geom_function(mapping = aes(colour = (sat$percent>=50)), fun = lm_above50_func)
sat %>%
  mutate(above50 = (percent>=50)) %>%
  ggplot(aes(x = percent, y = verbalScore)) +
  geom_point() +
  geom_function(mapping = aes(colour = above50), fun = lm_above50_func, size = 1)
```

Though the correct colouring for `geom_function` is lacking.
```{r}
sat %>% ggplot(aes(x = percent, y = verbalScore, colour = (percent>=50), group = 1)) +
  geom_point() +
  geom_line() +
  xlab('X') + ylab('Y') +
  scale_colour_manual(values = c("red", "blue")) +
  theme(legend.position='top')
```

```{r}
sat %>% ggplot(aes(x = percent, y = verbalScore, colour = (percent>=50), group = 1)) +
  geom_point() +
  geom_function(fun = lm_above50_func) +
  theme(legend.position = 'top')
```


## Question 3

Fit a natural cubic spline model with 4 knots. Use the `ns` function to generate a B-spline basis.
How should you choose the placements of the knots?
How can you formally test the linearity assumption within this larger model?

<!-- Standard "introduktion til statistik" side 133 svar / brug af anova -->



## Changing the output format

The default is to get an HTML file, which can be viewed in a browser. Occasionally, 
you might want to turn the result into pdf. This can be achieved in RStudio by selecting 
to knit to pdf. If you really want a nice report in pdf, it is 
better to integrate R with LaTeX instead of markdown.








